---
title: "Bioregion Delineation with ecoregionr: Unconstrained vs. Spatially Constrained Clustering"
author: "Mohamed Jumanne Omari"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: kate
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = TRUE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 7,
  dpi        = 150
)
```

---

# Introduction

## What is bioregionalization?

Bioregionalization — or ecological regionalization — is the process of dividing
a geographic area into zones that share similar environmental conditions. Rather
than using administrative boundaries (which are political artefacts), bioregions
are defined by measurable ecological signals: climate, elevation, vegetation
cover, soil moisture, species composition, and so on.

## Two strategies implemented in this package

The `ecoregionr` package provides **two complementary clustering strategies**,
each answering a slightly different question:

| Strategy | Core idea | Best for |
|---|---|---|
| **PCA + k-means** (unconstrained) | Group units that are *environmentally* most similar, ignoring geography | Maximising within-cluster homogeneity; hypothesis generation |
| **SKATER** (spatially constrained) | Group units that are environmentally similar *and* geographically connected | Operationally deployable contiguous regions; planning & logistics |

Neither is universally superior. Unconstrained clustering can reveal
environmentally coherent groups scattered across the map — useful for
understanding the environmental space. Spatially constrained clustering
sacrifices a little environmental purity to guarantee contiguous, navigable
zones — useful when field teams need to operate within a single continuous region.

This document walks through both methods on the same dataset, compares the
outputs, and explains when to prefer each approach.

---

# Setup and Configuration

## Install and load the package

```{r load-package, eval=FALSE}
# Install from GitHub (once)
# devtools::install_github("Medjumanne/ecoregionr")

# Load
library(ecoregionr)
```

```{r load-package-local, echo=FALSE}
# For local development — adjust path as needed
# devtools::load_all("~/Documents/ecoregionr")
library(ecoregionr)
```

## Required additional packages

```{r load-packages, message=FALSE}
packages <- c(
  "terra", "sf", "dplyr", "tidyr", "ggplot2",
  "rnaturalearth", "rnaturalearthdata",
  "leaflet", "htmlwidgets",
  "spdep",          # required for SKATER
  "factoextra",     # PCA visualisation
  "cluster",        # silhouette
  "patchwork",      # combine plots
  "ggspatial",      # north arrow + scale bar
  "viridis",
  "purrr"
)

for (pkg in packages) {
  if (!requireNamespace(pkg, quietly = TRUE))
    install.packages(pkg, dependencies = TRUE)
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}
```

## Configuration block

Adjust **only** this section between projects. Everything below reads from
these variables, so you do not need to edit the pipeline code.

```{r config}
# ── Paths ─────────────────────────────────────────────────────────────────────
WORKDIR   <- "~/Documents/VectorGrid/WP1_baselinbe_biogeographic_maps"
OUTDIR    <- file.path(WORKDIR, "output_TZ")
SHAPEFILE <- file.path(WORKDIR, "shepefiles_TZ/New Districtss.shp")

# ── Column names in the shapefile ─────────────────────────────────────────────
ID_COL   <- "ID"          # unique row identifier (created if absent)
UNIT_COL <- "NewDist18"   # human-readable polygon label (district name)

# ── Single-country mode ───────────────────────────────────────────────────────
# Set GROUP_COL = "ADM0_NAME" and COUNTRY_NAME = NULL for multi-country runs.
GROUP_COL    <- NULL
COUNTRY_NAME <- "United Republic of Tanzania"
STUDY_GROUPS <- NULL

# ── Continuous rasters ────────────────────────────────────────────────────────
CONTINUOUS_RASTERS <- list(
  elev         = file.path(WORKDIR, "output/elevation_vg_africa.tif"),
  pop          = file.path(WORKDIR, "output/population_vg_africa.tif"),
  urban        = file.path(WORKDIR, "output/urban_extent_vg_africa.tif"),
  transport    = file.path(WORKDIR, "output/transport_density_km_per_km2.tif"),
  mosquitoRich = file.path(WORKDIR, "output/mosquito_species_richness_smoothed_per_all_area.tif"),
  gambiae      = file.path(WORKDIR, "shepefiles_TZ/Mosq_abu/gambiae_clipped.tif"),
  funestus     = file.path(WORKDIR, "shepefiles_TZ/Mosq_abu/funestus_clipped.tif"),
  arabiensis   = file.path(WORKDIR, "shepefiles_TZ/Mosq_abu/arabiensis_clipped.tif")
)

CONTINUOUS_FUNS <- list(
  elev         = list(median = median),
  pop          = list(mean   = mean),
  urban        = list(mean   = mean),
  transport    = list(mean   = mean),
  mosquitoRich = list(max    = max),
  gambiae      = list(mean   = mean),
  funestus     = list(mean   = mean),
  arabiensis   = list(mean   = mean)
)

AGG_FACTS <- list(elev = 1, pop = 1, urban = 10, transport = 1, mosquitoRich = 1)

# ── Categorical rasters ────────────────────────────────────────────────────────
CATEGORICAL_RASTERS <- list(
  kg        = file.path(WORKDIR, "output/climate_vg_africa.tif"),
  landcover = file.path(WORKDIR, "output/lc_study_area_final.tif")
)

KG_LOOKUP       <- ecoregionr:::kg_lookup
LC_LOOKUP       <- ecoregionr:::lc_lookup
LC_EXCLUDE_VALS <- c(10, 4294967295, 4294967296)
LC_CLASS_NAMES  <- unname(LC_LOOKUP)

# ── Numeric variable detection ─────────────────────────────────────────────────
# NULL = auto-detect columns ending in _mean, _median, _max, _coverage, _rate
NUMERIC_VARS <- NULL

# ── Shared clustering parameters ──────────────────────────────────────────────
SEED <- 42
```

---

# Data Preparation (shared by both methods)

The steps in this section are identical for both clustering strategies.
We load the polygon layer, extract raster statistics, build a covariate table,
and impute missing values. The resulting `env_df` data frame feeds into both
methods.

## Step 1 — Load and validate polygons

```{r load-polygons}
adm2 <- sf::st_read(SHAPEFILE, quiet = TRUE)

# Add a sequential ID column if absent
if (!ID_COL %in% names(adm2)) adm2[[ID_COL]] <- seq_len(nrow(adm2))

# Single-country mode: inject a synthetic .group column
if (is.null(GROUP_COL)) {
  GROUP_COL         <- ".group"
  adm2[[GROUP_COL]] <- COUNTRY_NAME
}

if (!is.null(STUDY_GROUPS))
  adm2 <- adm2[adm2[[GROUP_COL]] %in% STUDY_GROUPS, ]

METADATA_COLS <- GROUP_COL
adm2 <- sf::st_make_valid(adm2)

cat("Polygons loaded:", nrow(adm2), "\n")
```

## Step 2 — Load rasters and reproject polygons

```{r load-rasters}
cont_rasters <- lapply(CONTINUOUS_RASTERS, terra::rast)
cat_rasters  <- lapply(CATEGORICAL_RASTERS, terra::rast)

ref_crs <- terra::crs(cont_rasters[[1]])
adm2    <- sf::st_transform(adm2, crs = ref_crs)
```

## Step 3 — Extract continuous raster statistics

For each polygon we summarise every raster with its designated function
(median for elevation, mean for population, max for mosquito richness, etc.).

```{r extract-continuous}
adm2_stats <- extract_continuous_stats(
  polygons       = adm2,
  rasters        = cont_rasters,
  funs           = CONTINUOUS_FUNS,
  id_col         = ID_COL,
  aggregate_fact = AGG_FACTS
)

# Round mosquito richness to integers (it is a species count)
adm2_stats$mosquitoRich_max <- round(adm2_stats$mosquitoRich_max)
```

## Step 4 — Protected area coverage (optional covariate)

We calculate the proportion of each district that falls within a protected area.
This variable captures a conservation dimension that influences both vector
habitat and human access to services.

```{r protected-areas, eval=FALSE}
# Load and merge WDPA layers (adjust paths as needed)
protected_combined <- dplyr::bind_rows(
  sf::st_read("./shepefiles_TZ/...shp_0/...-polygons.shp", quiet = TRUE),
  sf::st_read("./shepefiles_TZ/...shp_1/...-polygons.shp", quiet = TRUE),
  sf::st_read("./shepefiles_TZ/...shp_2/...-polygons.shp", quiet = TRUE)
) %>%
  sf::st_transform(crs = ref_crs) %>%
  sf::st_make_valid()

protected_dissolved <- sf::st_union(protected_combined) %>% sf::st_make_valid()

sf::sf_use_s2(FALSE)
adm2_stats <- sf::st_make_valid(adm2_stats)

adm2_stats <- adm2_stats %>%
  dplyr::mutate(
    protected_area_coverage = sapply(seq_len(nrow(.)), function(i) {
      g   <- sf::st_geometry(adm2_stats[i, ])
      int <- suppressWarnings(sf::st_intersection(g, protected_dissolved))
      if (length(int) == 0) return(0)
      as.numeric(sf::st_area(int) / sf::st_area(g))
    })
  )
```

## Step 5 — Extract climate and land-cover compositions

Each polygon is summarised as a vector of proportions across Köppen-Geiger
climate classes and ESRI land-cover classes.

```{r extract-categorical}
kg_wide <- extract_categorical_composition(
  polygons   = adm2,
  cat_raster = cat_rasters$kg,
  lookup     = KG_LOOKUP,
  id_col     = ID_COL
)

lc_wide <- extract_categorical_composition(
  polygons     = adm2,
  cat_raster   = cat_rasters$landcover,
  lookup       = LC_LOOKUP,
  exclude_vals = LC_EXCLUDE_VALS,
  id_col       = ID_COL
)
```

## Step 6 — Build the covariate table

We join continuous stats and categorical compositions into a single wide table
that serves as the input to both clustering methods.

```{r build-table}
# Auto-detect numeric variable names
if (is.null(NUMERIC_VARS)) {
  all_cols     <- names(sf::st_drop_geometry(adm2_stats))
  NUMERIC_VARS <- grep("_mean$|_median$|_max$|_coverage$|_rate$", all_cols, value = TRUE)
  message("Auto-detected NUMERIC_VARS: ", paste(NUMERIC_VARS, collapse = ", "))
}

env_df <- build_cluster_table(
  polygons   = adm2_stats,
  cat_tables = list(kg_wide, lc_wide),
  keep_cols  = c(METADATA_COLS, NUMERIC_VARS),
  id_col     = ID_COL
)

# Fill NA proportions with 0 (class absent from polygon)
cat_prop_cols <- setdiff(names(env_df), c(ID_COL, METADATA_COLS, NUMERIC_VARS))
env_df <- env_df %>%
  dplyr::mutate(dplyr::across(dplyr::all_of(cat_prop_cols),
                               ~ tidyr::replace_na(.x, 0)))

glimpse(env_df)
```

## Step 7 — Median imputation (within-group)

Missing values in continuous covariates are replaced with the within-group
(within-country) median. Imputing per-group rather than globally ensures that
each country's feature matrix is self-consistent regardless of which other
countries are included in the run.

```{r impute}
env_df <- env_df %>%
  dplyr::group_by(.data[[GROUP_COL]]) %>%
  dplyr::mutate(dplyr::across(
    dplyr::all_of(NUMERIC_VARS),
    ~ ifelse(is.na(.x), median(.x, na.rm = TRUE), .x)
  )) %>%
  dplyr::ungroup()

# Variables used in both clustering methods
vars_for_clustering <- c(
  NUMERIC_VARS,
  setdiff(names(kg_wide), ID_COL),
  setdiff(names(lc_wide), ID_COL)
)

# Attach env_df to the sf object for SKATER.
# We only join the categorical proportion columns (climate + land cover) since
# adm2_stats already contains the numeric covariates. Joining env_df directly
# would create duplicate columns (elev_median.x / elev_median.y etc.) and
# break the vars_for_clustering name lookup.
cat_only_cols <- c(ID_COL, setdiff(names(kg_wide), ID_COL), setdiff(names(lc_wide), ID_COL))
adm2_sf <- adm2_stats %>%
  dplyr::left_join(env_df[, cat_only_cols], by = ID_COL) %>%
  sf::st_as_sf()
```

---

# Method 1 — Unconstrained Clustering: PCA + k-means

## Conceptual background

PCA (Principal Component Analysis) reduces the high-dimensional covariate
space to a smaller set of orthogonal axes that capture the dominant gradients
of environmental variation. The first few PCs typically represent interpretable
gradients: PC1 might be an "aridity axis", PC2 an "elevation axis", and so on.

K-means clustering then partitions the PC space into *k* clusters by minimising
the total within-cluster sum of squares. The optimal *k* is selected via
consensus voting across multiple NbClust indices.

**Key property**: there is no spatial constraint — two districts on opposite
sides of the country can be assigned to the same cluster if they share similar
environmental conditions. This is the method's strength for scientific insight
but its limitation for operational planning.

## Elevation banding — an optional stratification

Elevation banding is a **soft, optional constraint** that splits the data into
elevation strata (lowland / midland / highland) and runs k-means independently
within each stratum.  Use it when you expect the key environmental gradients to
differ fundamentally between altitudinal zones — for example, malaria drivers
in coastal lowlands vs. highland forests are so different that pooling them in
one PCA can mask within-stratum structure.

**When to skip it:** If your study area is relatively flat, or you want a single
set of clusters that spans all elevations, set `USE_ELEV_BAND <- FALSE` below
and `constraint_mode = "none"` in `run_clustering()`.

```{r kmeans-options}
# ── Toggle elevation banding here ────────────────────────────────────────────
USE_ELEV_BAND <- TRUE    # set FALSE to run a single unconstrained clustering

MIN_K   <- 6
MAX_K   <- 15
PCA_VAR <- 0.90
NSTART  <- 50
```

```{r kmeans-elev-band}
if (USE_ELEV_BAND) {
  env_df <- env_df %>%
    dplyr::mutate(
      elev_band = dplyr::case_when(
        elev_median < 300 ~ "lowland",
        elev_median < 900 ~ "midland",
        TRUE              ~ "highland"
      ),
      elev_band = factor(elev_band,
                         levels  = c("lowland", "midland", "highland"),
                         ordered = TRUE)
    )
  message("Elevation banding ON — stratum counts:")
  print(table(env_df$elev_band))
} else {
  message("Elevation banding OFF — single clustering across all elevations.")
}
```

## Run PCA + k-means clustering

```{r run-kmeans, message=TRUE}
clusters_km <- run_clustering(
  df              = env_df,
  weights         = NULL,
  constraint_var  = if (USE_ELEV_BAND) "elev_band" else NULL,
  constraint_mode = if (USE_ELEV_BAND) "split"     else "none",
  group_col       = GROUP_COL,
  id_col          = ID_COL,
  min_k           = MIN_K,
  max_k           = MAX_K,
  pca_var         = PCA_VAR,
  nstart          = NSTART,
  visualize       = TRUE,
  seed            = SEED
)

# Summary: how many clusters (and strata, if banded)?
clusters_km %>%
  dplyr::group_by(.data[[GROUP_COL]]) %>%
  dplyr::summarise(n_units = dplyr::n(), k = dplyr::first(k_used), .groups = "drop")
```

## Cluster heatmap

The heatmap shows the mean z-score of each covariate within each cluster.
Variables with high positive z-scores (red) are the defining characteristics
of that cluster; those near zero (white) are unremarkable.

```{r kmeans-heatmap}
study_groups <- unique(clusters_km[[GROUP_COL]])

for (g in study_groups) {
  p <- make_cluster_heatmap(clusters_km, group_name = g, group_col = GROUP_COL)
  print(p)
}
```

## Join clusters to polygons and map

```{r kmeans-map}
kg_cols <- setdiff(names(kg_wide), ID_COL)
lc_cols <- setdiff(names(lc_wide), ID_COL)

env_summary_km <- env_df %>%
  dplyr::group_by(.data[[ID_COL]]) %>%
  dplyr::summarise(
    dplyr::across(dplyr::any_of(c(kg_cols, lc_cols)), ~ mean(.x, na.rm = TRUE)),
    .groups = "drop"
  )

adm2_km <- adm2_stats %>%
  dplyr::left_join(dplyr::select(clusters_km, dplyr::all_of(ID_COL), cluster), by = ID_COL) %>%
  dplyr::left_join(env_summary_km, by = ID_COL) %>%
  dplyr::filter(!is.na(cluster))

# Helper for popup text
wrap_popup <- function(vals, nms) {
  keep <- !is.na(vals) & vals > 0
  if (!any(keep)) return("None")
  vals <- vals[keep]; nms <- nms[keep]
  ord  <- order(vals, decreasing = TRUE)
  paste(paste0(nms[ord], ": ", round(vals[ord], 4)), collapse = "<br/>")
}

adm2_km <- adm2_km %>%
  dplyr::rowwise() %>%
  dplyr::mutate(
    kg_popup = wrap_popup(dplyr::c_across(dplyr::all_of(kg_cols)), kg_cols),
    lc_popup = wrap_popup(dplyr::c_across(dplyr::all_of(lc_cols)), lc_cols)
  ) %>%
  dplyr::ungroup()

# Country border for both maps
countries_sf <- rnaturalearth::ne_countries(
  country = if (!is.null(STUDY_GROUPS)) STUDY_GROUPS else COUNTRY_NAME,
  returnclass = "sf"
)

# Static map
map_km <- make_static_map(
  polygons_clustered = adm2_km,
  cluster_col        = "cluster",
  border_sf          = countries_sf,
  projection         = "+proj=moll",
  viridis_option     = "turbo",
  legend_title       = "Bioregion (k-means)",
  simplify           = TRUE,
  keep_prop          = 0.02
)
print(map_km)
```

## Interactive Leaflet map (k-means)

```{r kmeans-leaflet}
label_fn_km <- function(row) {
  htmltools::HTML(paste0(
    "<strong>", row[[UNIT_COL]], " (Cluster ", row$cluster, ")</strong><br/>",
    "<hr style='margin:4px 0;'>",
    "<strong>Elevation:</strong> ",        round(row$elev_median,    1), " m<br/>",
    "<strong>Population:</strong> ",       round(row$pop_mean,       1), " /km²<br/>",
    "<strong>Mosquito Richness:</strong> ", row$mosquitoRich_max,          "<br/>",
    "<strong>Urban:</strong> ",            sprintf("%.3f", row$urban_mean),    "<br/>",
    "<strong>Transport:</strong> ",        sprintf("%.3f", row$transport_mean), "<br/>",
    "<strong>Climate:</strong><br/>",      row$kg_popup,                  "<br/>",
    "<strong>Land cover:</strong><br/>",   row$lc_popup
  ))
}

intr_map_km <- build_leaflet_map(
  polygons_clustered = adm2_km,
  cluster_col        = "cluster",
  label_fn           = label_fn_km,
  border_sf          = countries_sf,
  palette_option     = "turbo",
  title              = "Bioregions (PCA + k-means)"
)

intr_map_km
```

---

# Method 2 — Spatially Constrained Clustering: SKATER

## Conceptual background

SKATER (Spatial 'K'luster Analysis by Tree Edge Removal) is an algorithm
proposed by Assunção *et al.* (2006) that enforces spatial contiguity as a
hard constraint: every cluster must form a geographically connected region.

The algorithm works in three stages:

1. **Spatial graph construction.** We define neighbours using a distance
   threshold (e.g. centroids within 45 km are neighbours). Isolated polygons
   (islands) are connected to their nearest mainland polygon.

2. **Minimum Spanning Tree (MST).** Edge weights in the adjacency graph are
   set to the Euclidean distance between polygons in PCA-score space. The MST
   is the spanning tree with the minimum total edge weight — it traces the
   most "environmentally similar" path through the adjacency graph.

3. **Tree pruning.** Removing the *k−1* edges with the highest cost
   (= greatest environmental dissimilarity) partitions the MST into *k*
   subtrees, each representing one bioregion.

**Key property**: Because the MST is built on the *spatial adjacency graph*,
every subtree (cluster) is by construction a connected region. The trade-off
is that two environmentally identical but geographically distant districts
cannot be merged — they must be in different clusters.

## SKATER options

The elevation boost (`elev_col` / `elev_weight`) is **optional**.  When
enabled, a scaled, multiplied copy of the elevation column is appended to the
PC-score matrix before the MST is built — this makes the algorithm prefer
cutting MST edges that cross large elevation changes, so highland and lowland
districts end up in different clusters.

**When to skip it:** If elevation is already well-captured by the first PCs
(check the loading plot after running), or if your clustering variables don't
include elevation, set `USE_ELEV_BOOST <- FALSE`.

```{r skater-options}
# ── Toggle elevation boost here ───────────────────────────────────────────────
USE_ELEV_BOOST    <- TRUE    # set FALSE to skip the elevation multiplier
ELEV_BOOST_COL    <- "elev_median"
ELEV_BOOST_WEIGHT <- 7       # multiplier applied to the scaled elev vector

N_CLUSTERS_SKATER <- 15      # adjust after inspecting the elbow plot below
```

## Selecting k with the elbow method

Before running SKATER we use an elbow plot to guide the choice of *k*. The
within-cluster SSE is computed in PC-score space for k = 1 … 20. We look for
the "elbow" — the point where SSE reduction per additional cluster starts to
flatten.

```{r skater-elbow, message=TRUE}
elbow_plot <- plot_skater_elbow(
  polygons_sf    = adm2_sf,
  vars           = vars_for_clustering,
  max_k          = 20,
  pca_var        = 0.95,
  elev_col       = if (USE_ELEV_BOOST) ELEV_BOOST_COL    else NULL,
  elev_weight    = if (USE_ELEV_BOOST) ELEV_BOOST_WEIGHT else 1,
  dist_threshold = 45000
)
```

Inspect the plot, then update `N_CLUSTERS_SKATER` in the options chunk above.

## Run SKATER clustering

```{r run-skater, message=TRUE}
adm2_sf <- run_skater_clustering(
  polygons_sf    = adm2_sf,
  vars           = vars_for_clustering,
  n_clusters     = N_CLUSTERS_SKATER,
  pca_var        = 0.95,
  elev_col       = if (USE_ELEV_BOOST) ELEV_BOOST_COL    else NULL,
  elev_weight    = if (USE_ELEV_BOOST) ELEV_BOOST_WEIGHT else 1,
  dist_threshold = 45000,
  k_fallback     = 1,
  seed           = SEED,
  cluster_col    = "skater_cluster"
)

table(adm2_sf$skater_cluster)
```

## PCA loading plot

This plot shows how the original covariates relate to PC1 and PC2 — the axes
used to weight the MST edges. Variables with long arrows in the direction of
PC1 are the primary drivers of the first environmental gradient.

```{r skater-pca-loadings}
plot_skater_pca_loadings(adm2_sf)
```

## Cluster heatmap

```{r skater-heatmap}
make_skater_heatmap(
  adm2_sf,
  cluster_col = "skater_cluster",
  title       = "Bioregion Characteristics (SKATER)",
  subtitle    = "Values are z-scores relative to the study-area mean"
)
```

## Static map

```{r skater-static-map}
# Custom palette for up to 16 clusters
ecoregion_cols <- c(
  "cyan", "#3CB44B", "#FFE119", "#4363D8", "#F58231",
  "#911EB4", "#42D4F4", "#F032E6", "#BFEF45", "#FABEBE",
  "#469990", "#E6BEFF", "#9A6324", "#800000", "#FFFAC8", "grey28"
)

map_skater <- ggplot2::ggplot(adm2_sf) +
  ggplot2::geom_sf(ggplot2::aes(fill = skater_cluster), colour = "white", linewidth = 0.05) +
  ggplot2::scale_fill_manual(values = ecoregion_cols, name = "Ecoregion") +
  ggspatial::annotation_scale(location = "bl", width_hint = 0.4, style = "ticks") +
  ggspatial::annotation_north_arrow(
    location = "bl", which_north = "true",
    pad_x = ggplot2::unit(0.2, "in"), pad_y = ggplot2::unit(0.4, "in"),
    style = ggspatial::north_arrow_fancy_orienteering()
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    panel.grid   = ggplot2::element_blank(),
    axis.text    = ggplot2::element_blank(),
    axis.ticks   = ggplot2::element_blank()
  ) +
  ggplot2::labs(
    title    = "Spatially Constrained Bioregions (SKATER)",
    subtitle = paste(N_CLUSTERS_SKATER, "contiguous clusters"),
    caption  = "Method: SKATER via spdep"
  )

print(map_skater)
```

## Interactive Leaflet map (SKATER)

```{r skater-leaflet}
# Build climate and land-cover popup strings using cur_data() (safe with spaces in names)
climate_cols   <- intersect(names(adm2_sf), unname(KG_LOOKUP))
landcover_cols <- intersect(names(adm2_sf), unname(LC_LOOKUP))

adm2_sf <- adm2_sf %>%
  dplyr::rowwise() %>%
  dplyr::mutate(
    kg_popup_sk = paste(
      Filter(Negate(is.null), lapply(climate_cols, function(x) {
        val <- dplyr::cur_data()[[x]]
        if (!is.na(val) && val > 0) paste0(x, ": ", sprintf("%.3f", val))
      })),
      collapse = "<br/>"
    ),
    lc_popup_sk = paste(
      Filter(Negate(is.null), lapply(landcover_cols, function(x) {
        val <- dplyr::cur_data()[[x]]
        if (!is.na(val) && val > 0) paste0(x, ": ", sprintf("%.3f", val))
      })),
      collapse = "<br/>"
    )
  ) %>%
  dplyr::ungroup()

label_fn_skater <- function(row) {
  htmltools::HTML(paste0(
    "<strong>", row[[UNIT_COL]], " (Cluster ", row$skater_cluster, ")</strong><br/>",
    "<hr style='margin:4px 0;'>",
    "<strong>Elevation:</strong> ",         round(row$elev_median,    1), " m<br/>",
    "<strong>Population:</strong> ",        round(row$pop_mean,       1), " /km²<br/>",
    "<strong>Mosquito Richness:</strong> ", row$mosquitoRich_max,          "<br/>",
    "<strong>Urban:</strong> ",             sprintf("%.3f", row$urban_mean),    "<br/>",
    "<strong>Transport:</strong> ",         sprintf("%.3f", row$transport_mean), "<br/>",
    "<strong>Climate:</strong><br/>",       row$kg_popup_sk,               "<br/>",
    "<strong>Land cover:</strong><br/>",    row$lc_popup_sk
  ))
}

intr_map_skater <- build_leaflet_map(
  polygons_clustered = adm2_sf,
  cluster_col        = "skater_cluster",
  label_fn           = label_fn_skater,
  border_sf          = countries_sf,
  palette_option     = ecoregion_cols,
  title              = "Bioregions (SKATER)"
)

intr_map_skater
```

---

# Method Comparison

## Side-by-side static maps

```{r compare-maps, fig.width=14, fig.height=6}
library(patchwork)

p_km <- map_km +
  ggplot2::labs(title = "A   PCA + k-means (unconstrained)",
                subtitle = "Maximises within-cluster homogeneity")

p_sk <- map_skater +
  ggplot2::labs(title = "B   SKATER (spatially constrained)",
                subtitle = "Guarantees contiguous bioregions")

p_km | p_sk
```

## When to use which method

```{r comparison-table, echo=FALSE}
comparison <- data.frame(
  Criterion = c(
    "Spatial contiguity of clusters",
    "Within-cluster environmental purity",
    "Sensitivity to isolated polygons",
    "Interpretability for field teams",
    "Computational cost",
    "Best for"
  ),
  `PCA + k-means` = c(
    "Not guaranteed",
    "Maximised",
    "Not applicable",
    "Moderate (clusters may be fragmented)",
    "Low",
    "Environmental hypothesis generation; multi-country comparisons"
  ),
  SKATER = c(
    "Guaranteed",
    "Slightly lower (contiguity costs some purity)",
    "Requires graph connectivity fixes (handled automatically)",
    "High (regions are navigable zones)",
    "Moderate (MST computation)",
    "Operational planning; surveillance zone design; logistics"
  ),
  check.names = FALSE
)

knitr::kable(comparison, align = "lll",
             caption = "Comparison of the two clustering strategies.")
```

---

# Saving Outputs

```{r save-outputs, eval=FALSE}
ensure_dir(OUTDIR)

# ── SKATER outputs ─────────────────────────────────────────────────────────────
ggsave(file.path(OUTDIR, "skater_map_static.png"),
       map_skater, width = 12, height = 10, dpi = 300)

ggsave(file.path(OUTDIR, "skater_elbow.png"),
       elbow_plot, width = 10, height = 7, dpi = 300)

htmlwidgets::saveWidget(
  intr_map_skater,
  file.path(OUTDIR, "skater_map_interactive.html"),
  selfcontained = FALSE
)

write.csv(
  sf::st_drop_geometry(adm2_sf[, c(UNIT_COL, "skater_cluster")]),
  file.path(OUTDIR, "skater_cluster_assignments.csv"),
  row.names = FALSE
)

# ── k-means outputs ────────────────────────────────────────────────────────────
ggsave(file.path(OUTDIR, "kmeans_map_static.png"),
       map_km, width = 12, height = 10, dpi = 300)

write.csv(
  sf::st_drop_geometry(adm2_km[, c(UNIT_COL, "cluster")]),
  file.path(OUTDIR, "kmeans_cluster_assignments.csv"),
  row.names = FALSE
)

message("All outputs saved to: ", OUTDIR)
```

---

# Session Information

```{r session-info}
sessionInfo()
```

---

# References

- Assunção, R. M., Neves, M. C., Câmara, G., & Freitas, C. D. C. (2006).
  Efficient regionalisation techniques for socio-economic geographical units using
  minimum spanning trees. *International Journal of Geographical Information
  Science*, 20(7), 797–811. <https://doi.org/10.1080/13658810600665111>

- Bivand, R. S., Pebesma, E., & Gómez-Rubio, V. (2013). *Applied Spatial Data
  Analysis with R* (2nd ed.). Springer. <https://asdar-book.org>

- Beck, H. E., Zimmermann, N. E., McVicar, T. R., Vergopolan, N., Berg, A., &
  Wood, E. F. (2018). Present and future Köppen-Geiger climate classification
  maps at 1-km resolution. *Scientific Data*, 5, 180214.

- Karra, K., et al. (2021). Global land use/land cover with Sentinel-2 and deep
  learning. *IGARSS 2021*. (ESRI 10m land cover product.)
